`mathml`: Translate R Expressions to MathML and LaTeX

### 2023-06-18 rj/mathml.Rmd
Review 1

---
title: "`mathml`: Translate R Expressions to MathML and LaTeX"
date: "2023-04-25"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the R\ Markdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(mathml)

mathout <- function(..., flags=NULL, env=parent.frame())
{
  if(knitr::is_html_output())
    return(mathml(..., flags=c(flags, list(cat=TRUE)), env=env))

  if(knitr::is_latex_output())
    return(mathjax(..., flags=c(flags, list(cat=TRUE)), env=env))

  warning("mathout: no knitr output specified")  
  mathjax(..., flags=c(flags, list(cat=TRUE)), env=env)
}

catmath <- function(term, ...)
{
  s <- substitute(term)
  mathout(canonical(s), ...)
}
```

@@

````{r}
library(mathml)
mathjax(term)
````

Some of the curly braces are not really needed in this simple example, but are
necessary in edge cases. We can include the output in a R\ Markdown document by
specifying `results='asis'` in the R\ code chunk of R\ Markdown, as is done in all
examples below. The R\ function `mathout()` has been defined in the preamble of
this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
LaTeX output.

````{r, results='asis'}
mathout(term)
````

@@

```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
mathout(term)

term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
mathout(term)
```

@@

```{r, results='asis'}
term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
mathout(term)

term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
mathout(term)
```

@@

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
mathout(term)

term <- quote(dot(a, b) + frac(1L, nodot(c, d + e)) + dfrac(1L, times(g, h)))
mathout(term)
```

@@

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
mathout(term)
```

@@

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
mathout(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
mathout(term)
```

@@

mathout(sgn)
mathout(call("<-", quote(sgn(x)), sgn))
```

@@

```{r, results='asis'}
term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
mathout(term)

term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
mathout(term)
```

@@

```{r, results='asis'}
term <- quote(integrate(sin, 0L, 2L*pi))
mathout(term)
eval(term)
```

@@

```{r, results='asis'}
mathout(canonical(term))
```

@@

```{r, results='asis'}
v <- 1:3
mathout(call("t", v))

A <- matrix(data=11:16, nrow=2, ncol=3)
B <- matrix(data=21:26, nrow=2, ncol=3)
term <- call("+", A, B)
mathout(term)
```

@@

```{r, results='asis'}
term <- quote(pbinom(successes, Ntotal, prob))
mathout(term)

hook(successes, k)
hook(quote(Ntotal), quote(N), quote=FALSE)
hook(prob, pi)
mathout(term)
```

@@

```{r, results='asis'}
hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
mathout(term)
```

@@

term <- quote(t <- dfrac(m_A - m_B, 
    sqrt(denote(s2_p, frac((n_A - 1L)*s2_A + (n_B - 1L)*s2_B, n_A + n_B - 2L),
                "the pooled variance.") * (frac(1L, n_A) + frac(1L, n_B)))))
mathout(term)
```

@@

```{r, results='asis'}
t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
mathout(t, flags=list(error="highlight"))
mathout(t, flags=list(error="fix"))
```

@@

For this purpose, the functions `mathml()` or `mathjax()` have an optional
argument `flags` which is a list with named elements. In this example, we use
this argument to tell \pkg{mathml} how to render such erroneous
expressions using the flag `error` which can be "asis", "highlight", "fix", 
or "ignore". For more examples, see Table \@ref(tab:mistakes).

@@

# display as E(D; mu), c is a scaling parameter
hook(ED_single(.C, .Mu), E(`;`(D, .Mu)))
mathout(call("=", quote(ED_single(c, mu)), ED_single))
```

@@

mathout(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
```

@@

hook(ED_async(.Tau, .C, .MA, .SA, .MB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB))))
mathout(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
```

@@

hook(ED(.Tau, .C, .MA, .SA, .MB, .SB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB, .SB))))
mathout(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
```

@@

hook(ET(.Tau, .C, .MA, .SA, .MB, .SB, .MM),
     E(`;`(T[.Tau], `,`(.MA, .SA, .MB, .SB, .MM))))
mathout(call("=", quote(E(T[tau])), ET))
```

@@

mathout(call("=", quote(z[tau]), z))
```

@@

mathout(call("=", quote(X["8 df"]^2L), gof))
```

@@

mathout(call("=", quote(hat(theta)), lsq))
```

@@

mathout(f1)
````

@@

mathout(f2)
````

@@

In its current version \pkg{mathml} has some limitations. For example, it is
currently not possible to use the functions of \pkg{mathml} for writing
inline formulas, here, the user has to adopt the usual LaTeX notation. The
reason is the missing chunk option results="asis" in inline output from
R\ Markdown, such that the HTML or LaTeX commands are quoted.
Moreover, for long equations, we did not yet find a convenient way to insert
line breaks. This is mostly due to lacking support by MathML and LaTeX
renderers. For example, in its current stage, the LaTeX package `breqn` [@breqn]
is mostly a proof of concept. Finally, \pkg{mathml} is one-way road, that is,
translation from LaTeX or HTML back to R is not 
possible [see @latex2r, for an example].

@@

term <- quote(lm(EOT ~ T0 + Therapy, data=d, na.action=na.fail))
mathout(term)
````

@@

term <- canonical(quote(nthroot(n=3L, 2L)))
mathout(term)
```

@@

term <- quote(nthroot(a * (b + c), 3L)^2L)
mathout(term)
term <- quote(a^(1L/3L) + a^{1L/3L} + a^(1.0/3L))
mathout(term)
```

>>>>

---
title: "`mathml`: Translate R Expressions to MathML and LaTeX"
date: "2023-06-17"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the R\ Markdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(mathml)
```

@@

````{r}
library(mathml)
substr(mathml(term), 1, 70)
mathjax(term)
````

Some of the curly braces are not really needed in the LaTeX output, but are
necessary in edge cases. The package also includes a function `mathout()` that
wraps a call to `mathml()` for HTML output and `mathjax()` for LaTeX output.
Moreover, the function `math(x)` adds the class `"math"` to its argument, such
that a special knitr printing function is invoked. We can include the output in
a R\ Markdown document by specifying `results="asis"` in the R\ code chunk of
R\ Markdown, as is done in all examples below.

````{r, results="asis"}
math(term)
````

Similarly, `inline()` produces inline output: `r inline(term)`.

@@

```{r, results='asis'}
term <- quote(1 + -2L + a + abc + "a" + phi + Phi + varphi + roof(b)[i, j]^2L)
math(term)

term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
math(term)
```

@@

```{r, results='asis'}
term <- quote(bold(b[x, 5L]) + bold(b[italic(x)]) + italic(ab) + italic(42L))
math(term)

term <- quote(tilde(a) + mean(X) + box(c) + cancel(d) + phantom(e) + prime(f))
math(term)
```

@@

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
math(term)

term <- quote(dot(a, b) + frac(1L, nodot(c, d + e)) + dfrac(1L, times(g, h)))
math(term)
```

@@

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
math(term)
term <- quote('%+-%'(mean(X), 1.96 * s / sqrt(N))) # functional form of '%+-%'
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})   # the same
math(term)
```

@@

```{r, results='asis'}
term <- quote(sin(x) + sin(x)^2L + cos(pi/2L) + tan(2L*pi) * expm1(x))
math(term)
term <- quote(choose(N, k) + abs(x) + sqrt(x) + floor(x) + exp(frac(x, y)))
math(term)
```

@@

math(sgn)
math(call("<-", quote(sgn(x)), sgn))
```

@@

```{r, results='asis'}
term <- quote(S[Y]^2L <- frac(1L, N) * sum(Y[i] - mean(Y))^2L)
math(term)

term <- quote(log(prod_over(L[i], i==1L, N)) <- sum_over(log(L[i]), i==1L, N))
math(term)
```

@@

```{r, results='asis'}
term <- quote(integrate(sin, 0L, 2L*pi))
math(term)
eval(term)
```

@@

```{r, results='asis'}
math(canonical(term))
```

@@

```{r, results='asis'}
v <- 1:3
math(call("t", v))

A <- matrix(data=11:16, nrow=2, ncol=3)
B <- matrix(data=21:26, nrow=2, ncol=3)
term <- call("+", A, B)
math(term)
```

@@

```{r, results='asis'}
term <- quote(pbinom(successes, Ntotal, prob))
math(term)

hook(successes, k)
hook(quote(Ntotal), quote(N), quote=FALSE)
hook(prob, pi)
math(term)
```

@@

```{r, results='asis'}
hook(pbinom(.K, .N, .P), sum_over(dbinom(i, .N, .P), i=0L, .K))
math(term)
```

@@

term <- quote(t <- dfrac(m_A - m_B, 
    sqrt(denote(s2_p, frac((n_A - 1L)*s2_A + (n_B - 1L)*s2_B, n_A + n_B - 2L),
                "the pooled variance.") * (frac(1L, n_A) + frac(1L, n_B)))))
math(term)
```

@@

```{r, results='asis'}
t <- quote(dfrac(omit_right(mean(D) - mu[0L]), s / sqrt(N)))
math(t, flags=list(error="highlight"))
math(t, flags=list(error="fix"))
```

@@

For this purpose, the functions `mathml()`, `mathjax()`, `mathout()`
and `math()` have an optional argument `flags` which is a list with named
elements. In this example, we use this argument to tell \pkg{mathml} how to
render such erroneous expressions using the flag `error` which can be "asis",
"highlight", "fix", or "ignore". For more examples,
see Table \@ref(tab:mistakes).

@@

# display as E(D; mu), c is a scaling parameter
hook(ED_single(.C, .Mu), E(`;`(D, .Mu)))
math(call("=", quote(ED_single(c, mu)), ED_single))
```

@@

math(call("=", quote(E(D["AB"])), quote(ED_single(c, mu_A + mu_B))))
```

@@

hook(ED_async(.Tau, .C, .MA, .SA, .MB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB))))
math(call("=", quote(ED_async(tau, c, mu_A, sigma_A, mu_B)), ED_async))
```

@@

hook(ED(.Tau, .C, .MA, .SA, .MB, .SB), E(`;`(D[.Tau], `,`(.MA, .SA, .MB, .SB))))
math(call("=", quote(ED(tau, c, mu_A, sigma_A, mu_B, sigma_B)), ED))
```

@@

hook(ET(.Tau, .C, .MA, .SA, .MB, .SB, .MM),
     E(`;`(T[.Tau], `,`(.MA, .SA, .MB, .SB, .MM))))
math(call("=", quote(E(T[tau])), ET))
```

@@

math(call("=", quote(z[tau]), z))
```

@@

math(call("=", quote(X["8 df"]^2L), gof))
```

@@

math(call("=", quote(hat(theta)), lsq))
```

@@

math(f1)
````

@@

math(f2)
````

@@

One limitation of the package is the lack of a convenient way to insert line
breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
For example, in its current stage, the LaTeX package `breqn` [@breqn] is mostly
a proof of concept. Moreover, \pkg{mathml} is one-way road, that is, translation
from LaTeX or HTML back to R is not possible [see @latex2r, for an example].

@@

term <- quote(lm(EOT ~ T0 + Therapy, data=d, na.action=na.fail))
math(term)
````

@@

term <- canonical(quote(nthroot(n=3L, 2L)))
math(term)
```

@@

term <- quote(nthroot(a * (b + c), 3L)^2L)
math(term)
term <- quote(a^(1L/3L) + a^{1L/3L} + a^(1.0/3L))
math(term)
```
